Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-10-28 analysis/paper/paper.Rmd
REVAMP PROJECT FOR {targets}
major revamp of everything to support a {targets} pipeline approach. Previous functionality and pipeline steps that were within Rmd / Scripts have been ported to specific functions within R/

```{r wxCovMap, fig.height = 5.5, fig.id="wxCovMap", fig.cap="Example spatial distribution of weather covariates across the study region for 05 May 2012. The NARR was used to predict weather covariates for each hour in a percent dry timeline record."}

load(file = here::here('data/cov_means_daily_2012.Rdata'))
grid_plot <- cov_means_daily_2012 %>% filter(year == 2012,
                           yday == 125) %>%
  inner_join(grid.sf, by = 'cell') %>%
  sf::st_as_sf()

world <- ne_countries(scale = "medium", returnclass = "sf") %>% 
  sf::st_transform(st_crs(grid_plot))

p_precip <- ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(grid_plot, aes(fill = precip), size = 0) +
  scale_fill_distiller(palette = "RdYlBu") +
  # scale_fill_viridis_c(option = "C") +
  labs(subtitle = "precipitation") +
  theme(legend.position = "none")

p_pressure <- ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(grid_plot, aes(fill = pressure), size = 0) +
  scale_fill_distiller(palette = "RdYlBu") +
  # scale_fill_viridis_c(option = "C") +
  labs(subtitle = "barometric pressure") +
  theme(legend.position = "none")

p_wind <- ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(grid_plot, aes(fill = wind), size = 0) +
  scale_fill_distiller(palette = "RdYlBu") +
  # scale_fill_viridis_c(option = "C") +
  labs(subtitle = "wind") +
  theme(legend.position = "none")

p_temp2 <- ggplot() +
  annotation_spatial(world, 
                     size = 0, 
                     fill = "grey70") +
  layer_spatial(grid_plot, aes(fill = temp2), size = 0) +
  scale_fill_distiller(palette = "RdYlBu") +
  # scale_fill_viridis_c(option = "C") +
  labs(subtitle = "temperature") +
  theme(legend.position = "none")

(p_precip + p_pressure) / (p_wind + p_temp2) +
  plot_annotation(
    title = 'Weather varies across the Bering and Chukchi seas',
    caption = "North American Regional Reanalysis (NARR), 2012 May 05")
```

@@

```{r runmodels}
if (params$run_models) {
source(here::here('analysis/scripts/model.R'))
} else {
  load(here::here('data/fit_ribbon.Rdata'))
  load(here::here('data/fit_bearded.Rdata'))
  load(here::here('data/fit_spotted.Rdata'))
}
```

@@

```{r runyearmodels}
if (params$run_models) {
 source(here::here('analysis/scripts/model_year_hf.R'))
 source(here::here('analysis/scripts/model_year_pl.R'))

} else {
 load(here::here('data/fit_ribbon_year.Rdata'))
 load(here::here('data/fit_spotted_year.Rdata'))
}
```

@@

```{r createnewdata}
# function to create a prediction data frame for a given age_sex group
create_newdata <- function(data, age_sex) {
  df_list <- vector(mode = "list", length = length({{age_sex}} ))

  for (a_s in {{age_sex}}) {
  range_yday <- data %>% filter(age_sex == a_s ) %>%
    summarize(start_day = min(yday),
              end_day = max(yday))
  start_day <- range_yday$start_day
  end_day <- range_yday$end_day
  n_days = (end_day - start_day) + 1

  # for wx covariates we'll use a gam to get values by day/hour since wx is likely
  # to vary w/in day over the season

  gam.baro <-
    gam(pressure ~ s(yday), data = data)
  gam.temp <-
    gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
  gam.wind <-
    gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
  gam.precip <-
    gam(precip ~ s(yday), data = data)

  wx_new_data <- data.frame(
    solar_hour = rep(0:23, each = n_days),
    yday = rep(start_day:end_day, times = 24)
  )

  temp_pred <- predict(gam.temp, newdata = wx_new_data)
  wind_pred <- predict(gam.wind, newdata = wx_new_data)
  baro_pred <- predict(gam.baro, newdata = wx_new_data)
  precip_pred <- predict(gam.precip, newdata = wx_new_data)

  df <- data.frame(
    age_sex = a_s,
    solar_hour = rep(0:23, each = n_days),
    yday = rep(start_day:end_day, times = 24),
    northing = mean(data$northing),
    temp2 = temp_pred,
    wind = wind_pred,
    pressure = baro_pred,
    precip = precip_pred
  ) %>%
    mutate(
      sin1 = sin(pi * solar_hour / 12),
      cos1 = cos(pi * solar_hour / 12),
      sin2 = sin(pi * solar_hour / 6),
      cos2 = cos(pi * solar_hour / 6),
      sin3 = sin(pi * solar_hour / 4),
      cos3 = sin(pi * solar_hour / 4),
    ) %>%
    mutate(day = (yday - 120) / 10,
           day2 = day ^ 2,
           day3 = day ^ 3)
  df_list[[a_s]] <- df
  }
  if(length({{age_sex}}) > 1) {
  df_out <- bind_rows(df_list) %>%
    mutate(age_sex = forcats::fct_relevel(
      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG OF YEAR"))
    )
  } else {
    df_out <- bind_rows(df_list)
  }
}
```

@@

```{r hfnewdata}
# get the ribbon seal fit object
load(here::here("data/fit_ribbon.Rdata"))
ribbon_newdata <- create_newdata(fit_ribbon$dataset,
                                 age_sex = levels(fit_ribbon$dataset$age_sex))

# create the model matrix
ribbon_mm <- model.matrix(fit_ribbon$fixed.formula[-2],
                data = ribbon_newdata)

#clean up extra intercept and get coef
fit_coef <- fit_ribbon$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

ribbon_newdata <- ribbon_newdata %>%
  mutate(logit_fits = as.vector(ribbon_mm %*% fit_coef),
         logit_fits_var = diag(ribbon_mm %*% fit_ribbon$covb %*% t(ribbon_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(ribbon_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
         )
```

@@

```{r type3}
ribbon_type3 <- tibble(
  fit_ribbon$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_type3 <- setNames(split(ribbon_type3, 
                               seq(nrow(ribbon_type3))), 
                         rownames(ribbon_type3))

spotted_type3 <- tibble(
  fit_spotted$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_type3 <- setNames(split(spotted_type3, 
                               seq(nrow(spotted_type3))), 
                         rownames(spotted_type3))

bearded_type3 <- tibble(
  fit_bearded$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
bearded_type3 <- setNames(split(bearded_type3, 
                               seq(nrow(bearded_type3))), 
                         rownames(bearded_type3))
```

@@

```{r plnewdata}
# get the spotted seal fit object
load(here::here("data/fit_spotted.Rdata"))

spotted_newdata <- create_newdata(fit_spotted$dataset,
                                 age_sex = levels(fit_spotted$dataset$age_sex))

# create the model matrix
spotted_mm <- model.matrix(fit_spotted$fixed.formula[-2],
                          data = spotted_newdata)

#clean up extra intercept and get coef
fit_coef <- fit_spotted$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

spotted_newdata <- spotted_newdata %>%
  mutate(logit_fits = as.vector(spotted_mm %*% fit_coef),
         logit_fits_var = diag(spotted_mm %*% fit_spotted$covb %*% t(spotted_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(spotted_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
  )
```

@@

```{r ebnewdata}
# get the bearded seal fit object
load(here::here("data/fit_bearded.Rdata"))

eb_data <- fit_bearded$dataset %>% 
  mutate(age_sex = forcats::as_factor("ALL AGES"))

bearded_newdata <- create_newdata(eb_data,
                                  age_sex = levels(eb_data$age_sex))

# create the model matrix
bearded_mm <- model.matrix(fit_bearded$fixed.formula[-2],
                           data = bearded_newdata)

#clean up extra intercept and get coef
fit_coef <- fit_bearded$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

bearded_newdata <- bearded_newdata %>%
  mutate(logit_fits = as.vector(bearded_mm %*% fit_coef),
         logit_fits_var = diag(bearded_mm %*% fit_bearded$covb %*% t(bearded_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(bearded_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
  )
```

@@

```{r createnewdatayr}
create_newdata_year <- function(data, age_sex) {
  df_list <- vector(mode = "list", length = length({{age_sex}} ))
  df_list_years <- vector(mode = "list", length = length(levels(data$year)))

  for (a_s in {{age_sex}}) {
  for (y in levels(data$year)) {
    d <- data %>% filter(year == y, age_sex == a_s)
    if(nrow(d) == 0) { next }
    range_yday <- d %>%
      summarize(start_day = min(yday),
                end_day = max(yday))
    if(range_yday$start_day > 152) {next}
    if(range_yday$end_day < 121) {next}
    print(range_yday)
    start_day <- range_yday$start_day
    end_day <- range_yday$end_day
    n_days = (end_day - start_day) + 1

    # for wx covariates we'll use a gam to get values by day/hour since wx is likely
    # to vary w/in day over the season

    gam.baro <-
      gam(pressure ~ s(yday), data = data)
    gam.temp <-
      gam(temp2 ~ s(yday) + s(as.numeric(solar_hour)), data = data)
    gam.wind <-
      gam(wind ~ s(yday) + s(as.numeric(solar_hour)), data = data)
    gam.precip <-
      gam(precip ~ s(yday), data = data)

    wx_new_data <- data.frame(
      solar_hour = 12,
      yday = start_day:end_day
    )

    temp_pred <- predict(gam.temp, newdata = wx_new_data)
    wind_pred <- predict(gam.wind, newdata = wx_new_data)
    baro_pred <- predict(gam.baro, newdata = wx_new_data)
    precip_pred <- predict(gam.precip, newdata = wx_new_data)

    df <- data.frame(
      age_sex = a_s,
      year = y,
      solar_hour = 12,
      yday = start_day:end_day,
      northing = mean(data$northing),
      temp2 = temp_pred,
      wind = wind_pred,
      pressure = baro_pred,
      precip = precip_pred
    ) %>%
      mutate(
        sin1 = sin(pi * solar_hour / 12),
        cos1 = cos(pi * solar_hour / 12),
        sin2 = sin(pi * solar_hour / 6),
        cos2 = cos(pi * solar_hour / 6),
        sin3 = sin(pi * solar_hour / 4),
        cos3 = sin(pi * solar_hour / 4),
      ) %>%
      mutate(day = (yday - 120) / 10,
             day2 = day ^ 2,
             day3 = day ^ 3)
    df_list_years[[y]] <- df
  }
    df_list[[a_s]] <- bind_rows(df_list_years)
  }
  df_out <- bind_rows(df_list) %>%
    mutate(age_sex = forcats::fct_relevel(
      age_sex,c("ADULT.F","ADULT.M","SUBADULT","YOUNG OF YEAR")),
      year = forcats::as_factor(as.numeric(year))
    )

  return(df_out)
}
```

```{r hfpvyear}
load(here::here('data/fit_ribbon_year.Rdata'))

ribbon_newdata_year <- create_newdata_year(fit_ribbon_year$dataset,
                                           age_sex = levels(fit_ribbon_year$dataset$age_sex))

# create the model matrix
ribbon_mm <- model.matrix(fit_ribbon_year$fixed.formula[-2],
                          data = ribbon_newdata_year)

#clean up extra intercept and get coef
fit_coef <- fit_ribbon_year$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

# multiply by the model coefficients and convert to probabilities with plogis
# (this came from J Laake code for Hood Canal estimates)
ribbon_newdata_year <- ribbon_newdata_year %>%
  mutate(logit_fits = as.vector(ribbon_mm %*% fit_coef),
         logit_fits_var = diag(ribbon_mm %*% fit_ribbon_year$covb %*% t(ribbon_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(ribbon_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
  )

load(here::here('data/fit_spotted_year.Rdata'))

spotted_newdata_year <- create_newdata_year(fit_spotted_year$dataset,
                                            age_sex = levels(fit_spotted_year$dataset$age_sex))

# create the model matrix
spotted_mm <- model.matrix(fit_spotted_year$fixed.formula[-2],
                           data = spotted_newdata_year)

#clean up extra intercept and get coef
fit_coef <- fit_spotted_year$fixed.effects %>%
  filter(!is.na(std.err)) %>%
  pull(estimate)

# multiply by the model coefficients and convert to probabilities with plogis
# (this came from J Laake code for Hood Canal estimates)
spotted_newdata_year <- spotted_newdata_year %>%
  mutate(logit_fits = as.vector(spotted_mm %*% fit_coef),
         logit_fits_var = diag(spotted_mm %*% fit_spotted_year$covb %*% t(spotted_mm)),
         logit_fits_lo95 = logit_fits - 1.96*sqrt(logit_fits_var),
         logit_fits_up95 = logit_fits + 1.96*sqrt(logit_fits_var),
         ho_prob = plogis(spotted_mm %*% fit_coef),
         lower95 = plogis(logit_fits_lo95),
         upper95 = plogis(logit_fits_up95)
  )
```

```{r type3_year}
ribbon_year_type3 <- tibble(
  fit_ribbon_year$typeIII.hypoth
) %>% 
  mutate(# Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_year_type3 <- setNames(split(ribbon_year_type3, 
                               seq(nrow(ribbon_year_type3))), 
                         rownames(ribbon_year_type3))

spotted_year_type3 <- tibble(
  fit_spotted_year$typeIII.hypoth
) %>% 
  mutate(# Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_year_type3 <- setNames(split(spotted_year_type3, 
                               seq(nrow(spotted_year_type3))), 
                         rownames(spotted_year_type3))
```

@@

```{r annualHO, fig.height = 5.5, include=TRUE, fig.id = "annualHO", fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age-sex + year) that included observations from more than one seal are shown. Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}
select_groups <- fit_ribbon_year$dataset %>%
  filter(age_sex != "YOUNG OF YEAR") %>%
  group_by(age_sex, year) %>%
  summarise(n = n_distinct(speno), group_id = paste0(age_sex,year)) %>%
  filter(n >1) %>% pull(group_id)

@@

select_groups <- fit_spotted_year$dataset %>%
  filter(age_sex != "YOUNG OF YEAR") %>%
  group_by(age_sex, year) %>%
  summarise(n = n_distinct(speno), group_id = paste0(age_sex,year)) %>%
  filter(n >1) %>% pull(group_id)

@@

plot_df <- plot_df %>% bind_rows(plot_df2)
peak_ho_df <- peak_ho_df %>% bind_rows(peak_ho_df2)

>>>>

@@

```{r modelfits}
tar_load(bearded_fit, store = here::here("_targets"))
tar_load(ribbon_fit, store = here::here("_targets"))
tar_load(spotted_fit, store = here::here("_targets"))
tar_load(spotted_year_fit, store = here::here("_targets"))
tar_load(ribbon_year_fit, store = here::here("_targets"))
```

@@

@@

@@

```{r hfnewdata}
targets::tar_load(ribbon_newdata, store = here::here("_targets"))
```

@@

```{r type3}
ribbon_type3 <- tibble(
  ribbon_fit$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_type3 <- setNames(split(ribbon_type3, 
                               seq(nrow(ribbon_type3))), 
                         rownames(ribbon_type3))

spotted_type3 <- tibble(
  spotted_fit$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_type3 <- setNames(split(spotted_type3, 
                               seq(nrow(spotted_type3))), 
                         rownames(spotted_type3))

bearded_type3 <- tibble(
  bearded_fit$typeIII.hypoth
) %>% 
  mutate(Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
bearded_type3 <- setNames(split(bearded_type3, 
                               seq(nrow(bearded_type3))), 
                         rownames(bearded_type3))
```

@@

```{r plnewdata}
targets::tar_load(spotted_newdata, store = here::here("_targets"))
```

@@

```{r ebnewdata}
targets::tar_load(bearded_newdata, store = here::here("_targets"))
```

@@

```{r hfpvyear}
targets::tar_load(ribbon_newdata_year, store = here::here("_targets"))
targets::tar_load(spotted_newdata_year, store = here::here("_targets"))
```

```{r type3_year}
ribbon_year_type3 <- tibble(
  ribbon_year_fit$typeIII.hypoth
) %>% 
  mutate(# Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>% 
  column_to_rownames(var = "Effect") 
ribbon_year_type3 <- setNames(split(ribbon_year_type3, 
                               seq(nrow(ribbon_year_type3))), 
                         rownames(ribbon_year_type3))

spotted_year_type3 <- tibble(
  spotted_year_fit$typeIII.hypoth
) %>% 
  mutate(# Prob.F = ifelse(Prob.F<0.001, "<0.01",Prob.F),
         Prob.F = scales::pvalue(Prob.F),
         Effect = janitor::make_clean_names(Effect)) %>%  
  column_to_rownames(var = "Effect")  
spotted_year_type3 <- setNames(split(spotted_year_type3, 
                               seq(nrow(spotted_year_type3))), 
                         rownames(spotted_year_type3))
```

@@

```{r annualHO, fig.height = 5.5, include=TRUE, fig.id = "annualHO", fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age-sex + year) that included observations from more than one seal are shown. Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}
select_groups <- ribbon_year_fit$dataset %>%
  filter(age_sex != "YOUNG OF YEAR") %>%
  group_by(age_sex, year) %>%
  summarise(n = n_distinct(speno), group_id = paste0(age_sex,year)) %>%
  filter(n >1) %>% pull(group_id)

@@

select_groups <- spotted_year_fit$dataset %>%
  filter(age_sex != "YOUNG OF YEAR") %>%
  group_by(age_sex, year) %>%
  summarise(n = n_distinct(speno), group_id = paste0(age_sex,year)) %>%
  filter(n >1) %>% pull(group_id)

@@

plot_df <- plot_df %>% 
  bind_rows(plot_df2) %>% 
  mutate(year = forcats::fct_relevel(.$year, sort))
peak_ho_df <- peak_ho_df %>% 
  bind_rows(peak_ho_df2) %>% ungroup() %>% 
  mutate(year = forcats::fct_relevel(.$year, sort))
